Generating Correlated Data for Omics Simulation

### 2025-05-27 manuscript/paper.qmd
Discuss CYCLOPS performance, revise SPsimSeq discussion

By default, CYCLOPS performs dimension reduction so that each dimension (called an "eigengene") contains at least 3% of the total variance.
We found that CYCLOPS performance depended significantly on this parameter, with the default producing good performance across all simulations.
However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
At that setting, CYCLOPS performance is much higher in the PCA method and moderately improved in Wishart method, compared to the independent method. 
This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
This demonstrates that the correlation structure of the transcriptome can have a major impact on performance.

@@

We compared our simulated data to that of the R package SPsimSeq [@Assefa2020].
In contrast to our package, SPsimSeq uses a two-step randomization.
In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference dataset.
These values are used to choose which bucket each gene is drawn from.
Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will depended upon parameters such as the number of buckets used.
More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference dataset.
SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
In comparison, our dependentsimr package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

>>>>

By default, CYCLOPS performs dimension reduction so that each dimension (called an "eigengene") contains at least 3% of the total variance.
We found that CYCLOPS performance depended significantly on this parameter, with the default producing good performance across all simulations.
However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
At that setting, CYCLOPS performance is much higher in the PCA method and moderately improved in Wishart method, compared to the independent method. 
This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
This demonstrates that the correlation structure of the transcriptome can have a major impact on performance.
In this application, the differences between corpcor and Wishart were such that, while Wishart improved upon the number of eigengenes used compared to corpcor, both undershot the level of correlation in the real data set (see Supplemental Figure 2 d) and the difference between them was too small to affect the actual performance of CYCLOPS as measured by the circular correlation.

@@

We compared our simulated data to that of the R package SPsimSeq [@Assefa2020].
In contrast to our package, SPsimSeq uses a two-step randomization.
In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference dataset.
These values are used to choose which bucket each gene is drawn from.
Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will depended upon parameters such as the number of buckets used (more fine-grained buckets will produce higher correlation).
SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
In comparison, our dependentsimr package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.
